---
title: "A3: Incarceration"
author: "Shipei Huang"
date: "08/04/23"
output: html_document
---
```{r setup, include=FALSE}
source("coding.R")
```

### Introduction
The U.S. has the most people in jail worldwide, and Black Americans, due to historical and current racism, are locked up much more than others. In this report, I chose the last data set (a smaller dataset about prison/jail population per 100,000 people (rate) per county/state starting at 1990 in Washington) given in our homework instruction. The reason I chose this dataset is because I live in WA state and would like to know more about the incarceration in our state.
                    
### Summary Information
5 Variables I calculated: 

1. 'average_pop': Calculated out the average value of population in jails of each year throughout all the WA counties.
    Max: 193220.3 in 2018.  Min: 125719.1 in 1990
    
2. 'average_jail_pop_rate of 1990 and 2018': Calculated out the average jail population rate at the earliest and most recent year.
    average_jail_pop_rate_1990 = 221.0728.  average_jail_pop_rate_2018 = 326.6744
    
3. 'median_jail_pop_rate':  Calculated out the median jail rate from 1990 to 2018 throughout all WA counties.
    The median observation is Whatcom County in 2007 with total_jail_pop_rate = 277.89

4. 'highest_jail_pop_rate': Calculated out the county with highest jail population rate throughout all WA counties in this dataset.
    Yakima County had the highest jail population rate 992.82 in 2007

5. 'average_black_rate vs. average_white_rate': Calculated out the average jail rate for black and white people throughout counties.
    average_black_rate = 1868.63, 	average_white_rate = 261.5801; Huge difference between white and black people!!

```{r}
average_pop
average_jail_pop_rate_1990
average_jail_pop_rate_2018
median_jail_pop_rate
highest_jail_pop_rate
average_black_rate
average_white_rate
```

### The Dataset

1. Who collected the data? 
This dataset was assembled using information collected by the U.S. Department of Justice Bureau of Justice Statistics (BJS), supplemented with data from state departments of correction when federal data is not available

2. How was the data collected or generated? 
The prison variables draw on the National Corrections Reporting Program (NCRP) and data collected directly from state departments of correction, when NCRP data is not available or is unreliable. These two sources are combined when creating the prison variables. The jails data draws on the Annual Survey of Jails (ASJ), Census of Jails (COJ), and population and admissions variables from Death in Custody Reporting Program (DCRP). The ASJ and COJ data are combined when creating the jail variables. This dataset provides county-level data on prison and jail incarceration and related measures over time for the entire United States.

3. Why was the data collected?  
The purpose of this dataset is to provide detail on the sources and variables in the Incarceration Trends dataset.This dataset is a good reference for important topics like racial and money-related differences, decisions made by the government, and fairness problems in the system.

4. How many observations (rows) are in your data?  
1131 rows

5. How many features (columns) are in the data?  
23columns

6. What, if any, ethical questions or questions of power do you need to consider when working with this data?  
I think we need to consider the fairness and bias problem that may exist when working with this data. If the data is not objective and itself has bias. Then our trending analysis would be biased as well. I would also consider the transparency of this data when working with it. For example, I will consider that is the data collection process transparent, and can others understand how the data was gathered? Is there a risk of hidden agendas or manipulation. 


7. What are possible limitations or problems with this data? 
Using this data the wrong way could accidentally support wrong ideas or keep unfair things going. The data comes mostly from the government or big organizations, which might make it not completely fair because they collect and understand it differently. Also, sometimes the data from these places might not be accurate because they want to look good or follow certain rules. 


### Trends Over Time Chart
```{r}
chart1
```

I chose King County because we live here and I chose Whatcom county because it's the median observation among all
the WA counties. As we can see from the graph, King County had a lower average black jail population from 1990 to 2018. It also had smaller fluctuations in jail population change from year to year in overall. Since Whatcom county was the median observation in our data set (based on the median_jail_pop_rate), we can believe that King county is a relatively safe place than most of the WA Counties.

### Variable Comparison Chart
```{r pressure, echo=FALSE}
chart2
```

This graph shows the weak relationship between average jail population rate of the black and white people. We can
still see that the black people have higher average jail population rate than white people. Based
on the graph above, there is a very weak relationship between white jail population and black
jail population. The distribution of the points is so random and the slope of the regression line is small.

Based on the regression line, we can predict that as white jail population rate increases, the black jail population rate also increases. This totally make sense especially when there is high unemployment.

### Map
```{r}
chart3
```

I chose this chart because I live in WA state all the time and want to figure out the most recent jail data for different counties. Based on the map above, we can see that most of the counties have total jail rate around 400 in 2018. In overall, they have similar jail rates (no big difference).

